Understanding the Risk of Casualties in US School Shootings

### 2024-12-03 paper/paper.qmd
update pictures and scripts

---
title: "2025 US School Shooting Predictions and Solutions"
subtitle: "Leveraging Forecasting to Shape Policy and Protect Communities"
author: 
  - Yun Chu
thanks: "Code and data are available at: https://github.com/chuyun2024/School-Shooting-Analysis."
date: today
date-format: long
abstract: "School shootings in the United States pose a critical challenge with far-reaching impacts. This paper employs a Random Forest regression model to predict state-level casualties for 2025 using historical data from 1999 to the present. The findings reveal a positively skewed distribution of casualties, with states like Georgia, California, and Pennsylvania projected to face the highest numbers. These insights highlight the need for targeted interventions, including enhanced school safety measures and stricter gun control policies. By providing actionable predictions, this study aims to inform policies that reduce school shooting casualties and protect vulnerable populations."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(ggplot2)
library(dplyr)
library(maps)
library(knitr)
library(modelsummary)
library(here)
library(kableExtra)
library(arrow)
#### Read data ####
data <- read_parquet("../data/02-analysis_data/analysis_data.parquet")
data$ifcasualty <- ifelse(data$casualties> 0, 1, 0)
# Create a new column with simplified categories
data$shooting_category <- ifelse(
  data$shooting_type %in% c("indiscriminate", "targeted"),  
  "indiscriminate or targeted",                              
  "other"
)

@@

The statistical programming language R [@citeR] is used to download, clean, analyze and model the US School Shooting Data. The US School Shooting dataset is downloaded from The Washington Post [@washingtonpost_school_shootings_data] . The following libraries are utilized in this paper: `tidyverse` [@tidyverse2024], `dplyr` [@dplyr2024], `lubridate` [@lubridate2024], `readr` [@readr2024], `stringr` [@stringr2024], `arrow` [@arrow2024], `testthat` [@testthat2024], `modelsummary` [@modelsummary], `ggplot2` [@ggplot2], `maps` [@maps], `knitr` [@knitr], `here` [@here], `kableExtra` [@kableExtra], `rstanarm` [@rstanarm].

@@

```{r}
#| label: fig-casualty-by-shooting-type
#| fig-cap: "Distribution of Casualty By Shooting Type"
#| echo: false
#| eval: true
#| 
# Load necessary libraries
library(ggplot2)
library(dplyr)

@@

# Plot the total number of casualties by shooting type
ggplot(casualties_by_type, aes(x = reorder(shooting_type, total_casualties), y = total_casualties, fill = shooting_type)) +
  geom_bar(stat = "identity", color = "white") +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Number of Casualties by Shooting Type",
    x = "Shooting Type",
    y = "Total Number of Casualties",
    fill = "Shooting Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.position = "none"  # Hide legend as labels are self-explanatory
  )
```

@@

```{r}
#| label: fig-casualtydist
#| echo: false
#| fig-cap: Distribution of Casualties
# Causaulity Distribution
ggplot(data, aes(x = casualties)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "white") +
  labs(x = "Number of Casualties", y = "Frequency") +
  theme_minimal()

@@

```{r}
#| label: fig-Shooting_Casaulties_by_State
#| fig-cap: "US School Shooting Casaulties by State"
#| echo: false

@@

# Plot the map
ggplot(map_data, aes(long, lat, group = group, fill = total_casualties)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(
    low = "#E6F7FF",  # Light blue
    high = "#0072B2",  # Dark blue
    name = "Total Casualties",
    breaks = c(0, 50, 100, 150, 200)  # Adjust breaks based on your data
  ) +
  labs(
    title = "US School Shooting Casualties by State",
    subtitle = "Number of casualties visualized geographically",
    x = NULL,
    y = NULL
  ) +
  coord_fixed(1.3) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = ggplot2::element_text(hjust = 0.5, face = "bold", size = 13, margin = ggplot2::margin(t = 10, b = 10)),
    plot.subtitle = ggplot2::element_text(hjust = 0.5, size = 10, margin = ggplot2::margin(t = 0, b = 20)),
    legend.position = "right",  # Alternative: Use "bottom" for a horizontal legend
    legend.key.height = unit(1, "cm"),  # Adjust legend size
    axis.text = ggplot2::element_blank(),
    axis.ticks = ggplot2::element_blank(),
    panel.grid = ggplot2::element_blank(),
    plot.margin = ggplot2::margin(t = 20, r = 20, b = 20, l = 20)  # Increase margins for PDF export
  )

@@

```{r}
#| label: fig-shootingtypedist
#| fig-cap: "Shooting Type Distribution" 
#| echo: false
#| eval: true

# Load necessary libraries
library(ggplot2)
library(dplyr)
library(ggrepel)

# Simulated dataset (replace this with your actual data)
# Assuming `data` has a column `shooting_type`
shooting_data <- data %>% 
  count(shooting_type) %>% 
  mutate(
    proportion = n / sum(n),  # Calculate proportions
    label = paste0(shooting_type, " (", scales::percent(proportion, accuracy = 1), ")")  # Labels
  )

# Calculate label positions and angles
shooting_data <- shooting_data %>%
  mutate(
    y_position = cumsum(proportion) - 0.5 * proportion,  # Calculate mid-point of each segment
    angle = 90 - (360 * (cumsum(proportion) - proportion / 2))  # Adjust angles for better label alignment
  )

# Plot Pie Chart
ggplot(shooting_data, aes(x = "", y = proportion, fill = shooting_type)) +
  geom_bar(width = 1, stat = "identity", color = "white") +
  coord_polar(theta = "y") +
  theme_void() +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Distribution of Shooting Types",
    fill = "Shooting Type"
  ) +
  geom_text(
    aes(
      label = ifelse(proportion > 0.05, paste0(shooting_type, " (", scales::percent(proportion, accuracy = 1), ")"), ""),
      y = y_position,
      angle = ifelse(proportion < 0.1, angle, 0)  # Add dynamic rotation for small segments
    ),
    size = 3,
    hjust = ifelse(shooting_data$proportion < 0.1, 1, 0.5),  # Align labels for smaller slices
    nudge_x = 0.4  # Nudge outward for better alignment
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.title = element_text(size = 10),
    legend.position = "right"
  )
```

@@

```{r}
#| label: fig-latitudedist
#| fig-cap: "Distribution of Latitude of Schools in School Shooting Events"
#| echo: false
#| eval: true
#| warning: false

# Load necessary libraries
library(ggplot2)

# Assuming `data` contains the `lat` column (latitude values)
# Replace `data` with your actual dataset

# Plot the distribution of latitude
ggplot(data, aes(x = lat)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Latitude",
    x = "Latitude",
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12)
  )

@@

## Intercept  
The intercept estimate is **2.6798** (95% credible interval: **1.0774 to 4.3091**), representing the baseline log-odds of a casualty occurring when predictors are at their reference levels. This corresponds to a high baseline probability, with the credible interval excluding zero, indicating strong evidence for a nonzero probability of casualties under these conditions.

## Shooting Type  
The coefficient for the "other" shooting category is **-0.6336** (95% credible interval: **-1.0735 to -0.1784**), suggesting a lower likelihood of casualties compared to the "indiscriminate" or "targeted" categories.

## Latitude  
The latitude coefficient is **-0.0443** (95% credible interval: **-0.0876 to -0.0021**), indicating a slight but significant decrease in the probability of casualties with increasing latitude. This suggests potential geographic trends influencing casualty outcomes, such as differences in population density, response capabilities, or regional policies.

@@

Using a Bayesian Logistic regression model, predictions for the existence of casualties in each state were generated. The predicted casualties were visualized geographically using a choropleth map in @fig-predicted_shooting_visualization.

```{r}
#| echo: false
#| eval: true
#| label: fig-predicted_shooting_visualization
#| fig-cap: "Existence of Casualties in US School Shootings Prediction"
#| warning: false
#| width: 8
#| height: 4
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(maps)
library(patchwork)
library(rstanarm)

us_states <- map_data("state")

# Calculate state centroids (mean latitude and longitude for each state)
state_centroids <- us_states %>%
  group_by(region) %>%
  summarise(long = mean(range(long)), lat = mean(range(lat)))

# Fix Florida's centroid directly
state_centroids <- state_centroids %>%
  mutate(
    long = ifelse(region == "florida", -81.686783, long),
    lat = ifelse(region == "florida", 27.766279, lat)
  )

# Continue with the rest of your code to predict and plot as before
state_data <- state_centroids %>%
  mutate(
    shooting_category = "indiscriminate or targeted",
    lat = lat,
    long = long
  )

# Predict probabilities and classify shooting existence
state_data$prob_targeted <- posterior_epred(bayes_model, newdata = state_data) %>% colMeans()
state_data$shooting_category <- "other"
state_data$prob_other <- posterior_epred(bayes_model, newdata = state_data) %>% colMeans()
state_data$exist_targeted <- ifelse(state_data$prob_targeted > 0.70657, 1, 0)
state_data$exist_other <- ifelse(state_data$prob_other > 0.70657, 1, 0)

# Plot for "Targeted/Indiscriminate"
p1 <- ggplot(us_states) +
  geom_polygon(aes(x = long, y = lat, group = group), fill = "lightgray", color = "white") +
  geom_point(data = state_data %>% filter(exist_targeted == 1), 
             aes(x = long, y = lat), 
             color = "darkred", size = 4, alpha = 0.8) +
  theme_void() +
  labs(title = "Shooting Type: Targeted/Indiscriminate", x = NULL, y = NULL) +
  theme(
    plot.title = element_text(size = 11, hjust = 0.5) # Shrink and center the title
  )

# Plot for "Other"
p2 <- ggplot(us_states) +
  geom_polygon(aes(x = long, y = lat, group = group), fill = "lightgray", color = "white") +
  geom_point(data = state_data %>% filter(exist_other == 1), 
             aes(x = long, y = lat), 
             color = "darkblue", size = 4, alpha = 0.8) +
  theme_void() +
  labs(title = "Shooting Type: Other", x = NULL, y = NULL) +
  theme(
    plot.title = element_text(size = 11, hjust = 0.5) # Shrink and center the title
  )

# Combine plots with patchwork
combined_plot <- p1 + p2 +
  plot_annotation(
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
  )

# Display the final plot
print(combined_plot)
```

@@

## **Analysis of The Washington Post's Data Collection Approach**

@@

### **Data Collection and Triangulation**
The Post utilized multiple data sources—including Nexis, news articles, open-source databases, law enforcement reports, and direct communications with schools and police departments—to compile a comprehensive list of school shootings. This method reflects the principle of data triangulation, which enhances the validity of findings by cross-verifying information from diverse sources [@denzin1978research]. Such triangulation is crucial in data journalism to ensure accuracy and comprehensiveness.

### **Defining Inclusion and Exclusion Criteria**
By establishing clear criteria for what constitutes a school shooting—excluding incidents like after-hours events, accidental discharges without injury, and private suicides—the Post ensured a focused analysis. This practice aligns with the concept of construct validity, ensuring that the study accurately reflects the phenomenon under investigation [@cronbach1955construct].

### **Sampling and Estimation Techniques**
To estimate the number of children exposed to gunfire, the Post used enrollment figures from the U.S. Education Department and adjusted for average daily absenteeism. Specifically, they deducted 7% from enrollment totals to account for average absenteeism and reduced enrollment by 50% for shootings occurring just before or after school hours. This approach mirrors sampling techniques discussed in "Telling Stories with Data", which emphasizes the importance of identifying the target population, understanding the context-specific variables (such as absenteeism rates), and applying appropriate adjustments to ensure the representativeness and reliability of the results [@alexander2023telling]. For example, the text highlights the necessity of incorporating real-world constraints—such as data availability or measurement gaps—into the design of sampling frameworks to improve the accuracy of population estimates.

### **Reproducibility and Transparency**
The Post's transparency in detailing their methodology allows for reproducibility and critical assessment, key components in data journalism. As highlighted in *Telling Stories with Data*, reproducible workflows involve sharing data, code, and analytical steps, enabling others to replicate and validate findings [@alexander2023telling].

### **Addressing Potential Biases**
The Post acknowledged that their database might not include every qualifying incident due to the absence of a federal tracking system. This recognition of potential data availability bias is essential, as discussed in the literature on observational data, which emphasizes the importance of considering who is included in the dataset and who is systematically excluded [@kaplan2014bias].

#### **Weaknesses**
Despite its strengths, The Washington Post’s methodology has notable limitations. Reliance on media reports and open-source databases may introduce reporting bias, potentially underrepresenting incidents in areas with limited media coverage [@kaplan2014bias]. The exclusion of certain incidents, such as private suicides, could understate the broader psychological impact of gun violence. Additionally, adjustments for absenteeism rates assume consistent patterns across all schools, which may overlook localized variations and specific community contexts [@alexander2023telling]. The use of proprietary tools like Nexis and reliance on non-public datasets also pose challenges for reproducibility, limiting the ability of others to independently verify findings. Finally, the absence of a centralized federal tracking system for school shootings creates a systemic data gap that even the Post’s robust methodology cannot fully address [@kaplan2014bias].

### **Conclusion**
The Washington Post's methodology in investigating school shootings demonstrates a rigorous application of data collection and analysis principles, aligning with best practices in data journalism and social science research. By employing data triangulation [@denzin1978research], defining clear inclusion criteria [@cronbach1955construct], adjusting for sampling considerations [@alexander2023telling], ensuring reproducibility [@alexander2023telling], and acknowledging potential biases [@kaplan2014bias], the Post provides a robust framework for understanding the impact of school shootings on children. Nonetheless, the identified weaknesses underscore the importance of continued refinement and broader systemic solutions for data collection.

>>>>

---
title: "Understanding the Risk of Casualties in US School Shootings"
subtitle: "Leveraging Bayesian Analysis to Guide Safety and Prevention Strategies"
author: 
  - Yun Chu
thanks: "Code and data are available at: https://github.com/chuyun2024/School-Shooting-Analysis."
date: today
date-format: long
abstract: "School shootings in the United States are a persistent and devastating issue. This paper utilizes a Bayesian Logistic Regression model to predict the probability of casualties in school shooting incidents, using historical data from 1999 onward. Key predictors, including shooting type and geographic latitude, are identified and analyzed. The results demonstrate that targeted and indiscriminate shootings significantly increase casualty risks, highlighting the need for targeted safety measures, such as enhanced threat assessment programs and stricter firearm regulations. This analysis provides a framework for implementing data-driven strategies to reduce casualties and improve school safety nationwide."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

library(tidyverse)
library(ggplot2)
library(dplyr)
library(maps)
library(knitr)
library(modelsummary)
library(here)
library(kableExtra)
library(arrow)
library(patchwork)
library(rstanarm)
#### Read data ####
data <- read_parquet("../data/02-analysis_data/analysis_data.parquet")
```

@@

The statistical programming language R [@citeR] is used to download, clean, analyze and model the US School Shooting Data. The US School Shooting dataset is downloaded from The Washington Post [@washingtonpost_school_shootings_data] . The following libraries are utilized in this paper: `tidyverse` [@tidyverse2024], `dplyr` [@dplyr2024], `lubridate` [@lubridate2024], `readr` [@readr2024], `stringr` [@stringr2024], `arrow` [@arrow2024], `testthat` [@testthat2024], `modelsummary` [@modelsummary], `ggplot2` [@ggplot2], `maps` [@maps], `knitr` [@knitr], `here` [@here], `kableExtra` [@kableExtra], `rstanarm` [@rstanarm], `arrow` [@R-arrow] and `patchwork` [@patchwork].

@@

```{r}
#| label: fig-casualty-by-shooting-type
#| fig-cap: "Number of Casualties for Each Shooting Type in US School Shooting Incidents"
#| echo: false
#| eval: true
#| 
# Load necessary libraries
library(ggplot2)
library(dplyr)

@@

# Plot the total number of casualties by shooting type
ggplot(casualties_by_type, aes(x = reorder(shooting_type, total_casualties), y = total_casualties, fill = shooting_type)) +
  geom_bar(stat = "identity", color = "white") +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Distribution of Casualty By Shooting Type",
    x = "Shooting Type",
    y = "Total Number of Casualties",
    fill = "Shooting Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),  # Rotate x-axis labels for readability
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    legend.position = "none"  # Hide legend as labels are self-explanatory
  )
```

@@

```{r}
#| label: fig-casualtydist
#| echo: false
#| fig-cap: "Distribution of Casualties in US School Shooting Incidents from 1999 to 2024"
# Causaulity Distribution
ggplot(data, aes(x = casualties)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "white") + 
  labs(title = "Distribution of Casualties", x = "Number of Casualties", y = "Frequency") +
  theme_minimal() +
    theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )

@@

```{r}
#| label: fig-Shooting_Casaulties_by_State
#| fig-cap: "US School Shooting Casaulties by State from 1999 to 2024"
#| echo: false

@@

# Adjusted Plot
ggplot(map_data, aes(long, lat, group = group, fill = total_casualties)) +
  geom_polygon(color = "white") +
  scale_fill_gradient(
    low = "#E6F7FF",
    high = "#0072B2",
    name = "Total Casualties",
    breaks = c(0, 50, 100, 150, 200)  # Adjust breaks based on your data
  ) +
  labs(
    title = "US School Shooting Casualties Visualization",
    x = NULL,
    y = NULL
  ) +
  coord_fixed(1.3) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14, margin = margin(t = 10, b = 10)),
    plot.subtitle = element_text(hjust = 0.5, size = 12, margin = margin(t = 0, b = 20)),
    legend.position = "right",
    legend.title = element_text(size = 10),  # Reduce the size of "Total Casualties"
    legend.text = element_text(size = 8),   # Reduce size of legend values
    legend.key.height = unit(0.5, "cm"),    # Reduce the height of the legend color bar
    legend.key.width = unit(0.25, "cm"),    # Reduce the width of the legend color bar
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 10)  # Increase margins for spacing
  ) +
  guides(
    fill = guide_colorbar(
      barwidth = 0.5,  # Adjust the width of the legend bar
      barheight = 8    # Adjust the height of the legend bar
    )
  )

@@

```{r}
#| label: fig-shootingtypedist
#| fig-cap: "US School Shooting Type Distribution from 1999 to 2024" 
#| echo: false
#| eval: true

# Simulated dataset (replace this with your actual data)
# Assuming `data` has a column `shooting_type`
shooting_data <- data %>% 
  count(shooting_type) %>% 
  mutate(
    proportion = n / sum(n)  # Calculate proportions
  )

# Plot Pie Chart without Text
ggplot(shooting_data, aes(x = "", y = proportion, fill = shooting_type)) +
  geom_bar(width = 1, stat = "identity", color = "white") +
  coord_polar(theta = "y") +
  theme_void() +
  scale_fill_brewer(palette = "Set3") +
  labs(
    title = "Distribution of Shooting Types",
    fill = "Shooting Type"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.title = element_text(size = 10),
    legend.position = "right"
  )

@@

```{r}
#| label: fig-latitudedist
#| fig-cap: "Distribution of Latitude of Schools in School Shooting Incidents in US from 1999 to 2024"
#| echo: false
#| eval: true
#| warning: false

# Plot the distribution of latitude
ggplot(data, aes(x = lat)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "white") +
  theme_minimal() +
  labs(
    title = "Distribution of Latitude of Schools in Shooting Incidents",
    x = "Latitude",
    y = "Frequency"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12)
  )

@@

## Intercept  
The intercept estimate is 2.37 (95% credible interval: 0.824 to 4.02), representing the baseline log-odds of a casualty occurring when predictors are at their reference levels. This corresponds to a high baseline probability, with the credible interval excluding zero, indicating strong evidence for a nonzero probability of casualties under these conditions.

## Shooting Type  
The coefficient for the "other" shooting category is -0.63 (95% credible interval: -1.09 to -0.16), suggesting a lower likelihood of casualties compared to the "indiscriminate" or "targeted" categories.

## Latitude  
The latitude coefficient is -0.04 (95% credible interval: -0.08 to 0.00), indicating a slight but significant decrease in the probability of casualties with increasing latitude. This suggests potential geographic trends influencing casualty outcomes, such as differences in population density, response capabilities, or regional policies.

@@

Using a Bayesian Logistic regression model, predictions for the existence of casualties in each state were generated. The predicted casualties were visualized geographically using a choropleth map in @fig-predicted_shooting_visualization where the states in dark red are classified as casualty exists while the states in grey are classified as no casualty.

```{r}
#| echo: false
#| eval: true
#| label: fig-predicted_shooting_visualization
#| fig-cap: "Prediction of the Existence of Casualty in US School Shootings Incidents Based on Bayesian Logistic Regression Model"
#| warning: false

# Load US map data
us_states <- map_data("state")

# Prepare state-level data for predictions
state_data <- us_states %>%
  distinct(region) %>%  # Get unique state names
  mutate(
    shooting_category = "indiscriminate or targeted" # Initial shooting category for predictions
  )

# Add latitude for each state by calculating the mean latitude of the state polygons
state_data <- state_data %>%
  left_join(
    us_states %>%
      group_by(region) %>%
      summarise(lat = mean(lat)),  # Calculate mean latitude for each state
    by = "region"
  )

# Predict probabilities using the Bayesian model
state_data$prob_targeted <- posterior_epred(bayes_model, newdata = state_data) %>% colMeans()
state_data$shooting_category <- "other"  # Change category for next prediction
state_data$prob_other <- posterior_epred(bayes_model, newdata = state_data) %>% colMeans()

# Classify shooting existence based on thresholds
state_data <- state_data %>%
  mutate(
    exist_targeted = ifelse(prob_targeted > 0.70657, 1, 0),
    exist_other = ifelse(prob_other > 0.70657, 1, 0)
  )

# Join predictions with the US map data
map_data <- us_states %>%
  left_join(state_data, by = "region") %>%  # Merge predictions with state polygons
  rename(lat = lat.x)  # Rename lat.x back to lat for compatibility with ggplot

# Plot for "Targeted/Indiscriminate"
p1 <- ggplot(map_data) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = factor(exist_targeted)), color = "white") +
  scale_fill_manual(values = c("0" = "lightgray", "1" = "darkred"), na.translate = FALSE) +
  coord_fixed(1.3) +  # Fix aspect ratio
  theme_void() +
  labs(
    title = "Shooting Type: Targeted/Indiscriminate",
    fill = NULL
  ) +
  theme(
    plot.title = element_text(size = 9, hjust = 0.5),
    plot.margin = margin(t = 5, r = 10, b = 5, l = 10)  # Add margins for spacing
  )

# Plot for "Other"
p2 <- ggplot(map_data) +
  geom_polygon(aes(x = long, y = lat, group = group, fill = factor(exist_other)), color = "white") +
  scale_fill_manual(values = c("0" = "lightgray", "1" = "darkblue"), na.translate = FALSE) +
  coord_fixed(1.3) +  # Fix aspect ratio
  theme_void() +
  labs(
    title = "Shooting Type: Other",
    fill = NULL
  ) +
  theme(
    plot.title = element_text(size = 9, hjust = 0.5),
    plot.margin = margin(t = 5, r = 10, b = 5, l = 10)  # Add margins for spacing
  )

# Combine the two plots using patchwork and add a title
combined_plot <- p1 + p2 +
  plot_annotation(
    title = "Predicted Existence of Casualties Visualization",
    theme = theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
      plot.margin = margin(t = 5, r = 5, b = 5, l = 5)  # Reduce margins around the combined plot
    )
  )

# Display the combined plot
print(combined_plot)

@@

## Analysis of The Washington Post's Data Collection Approach

@@

### Data Collection and Triangulation
The Post utilized multiple data sources—including Nexis, news articles, open-source databases, law enforcement reports, and direct communications with schools and police departments—to compile a comprehensive list of school shootings. This method reflects the principle of data triangulation, which enhances the validity of findings by cross-verifying information from diverse sources [@denzin1978research]. Such triangulation is crucial in data journalism to ensure accuracy and comprehensiveness.

### Defining Inclusion and Exclusion Criteria
By establishing clear criteria for what constitutes a school shooting—excluding incidents like after-hours events, accidental discharges without injury, and private suicides—the Post ensured a focused analysis. This practice aligns with the concept of construct validity, ensuring that the study accurately reflects the phenomenon under investigation [@cronbach1955construct].

### Sampling and Estimation Techniques
To estimate the number of children exposed to gunfire, the Post used enrollment figures from the U.S. Education Department and adjusted for average daily absenteeism. Specifically, they deducted 7% from enrollment totals to account for average absenteeism and reduced enrollment by 50% for shootings occurring just before or after school hours. This approach mirrors sampling techniques discussed in "Telling Stories with Data", which emphasizes the importance of identifying the target population, understanding the context-specific variables (such as absenteeism rates), and applying appropriate adjustments to ensure the representativeness and reliability of the results [@alexander2023telling]. For example, the text highlights the necessity of incorporating real-world constraints—such as data availability or measurement gaps—into the design of sampling frameworks to improve the accuracy of population estimates.

### Reproducibility and Transparency
The Post's transparency in detailing their methodology allows for reproducibility and critical assessment, key components in data journalism. As highlighted in *Telling Stories with Data*, reproducible workflows involve sharing data, code, and analytical steps, enabling others to replicate and validate findings [@alexander2023telling].

### Addressing Potential Biases
The Post acknowledged that their database might not include every qualifying incident due to the absence of a federal tracking system. This recognition of potential data availability bias is essential, as discussed in the literature on observational data, which emphasizes the importance of considering who is included in the dataset and who is systematically excluded [@kaplan2014bias].

### Weaknesses
Despite its strengths, The Washington Post’s methodology has notable limitations. Reliance on media reports and open-source databases may introduce reporting bias, potentially underrepresenting incidents in areas with limited media coverage [@kaplan2014bias]. The exclusion of certain incidents, such as private suicides, could understate the broader psychological impact of gun violence. Additionally, adjustments for absenteeism rates assume consistent patterns across all schools, which may overlook localized variations and specific community contexts [@alexander2023telling]. The use of proprietary tools like Nexis and reliance on non-public datasets also pose challenges for reproducibility, limiting the ability of others to independently verify findings. Finally, the absence of a centralized federal tracking system for school shootings creates a systemic data gap that even the Post’s robust methodology cannot fully address [@kaplan2014bias].

### Conclusion
The Washington Post's methodology in investigating school shootings demonstrates a rigorous application of data collection and analysis principles, aligning with best practices in data journalism and social science research. By employing data triangulation [@denzin1978research], defining clear inclusion criteria [@cronbach1955construct], adjusting for sampling considerations [@alexander2023telling], ensuring reproducibility [@alexander2023telling], and acknowledging potential biases [@kaplan2014bias], the Post provides a robust framework for understanding the impact of school shootings on children. Nonetheless, the identified weaknesses underscore the importance of continued refinement and broader systemic solutions for data collection.
